Classical mechanics as the high-entropy limit \\ of quantum mechanics

### 2025-12-02 Articles/2024-ClassicalLimit/classicalLimitArticle.tex
ClassicalLimit: fixed figure for landscape

In both classical and quantum mechanics, Gaussian states maximize entropy for a single independent degree of freedom at a fixed uncertainty or, equivalently, minimize uncertainty at fixed entropy. This means that, if we fix the entropy $S$, we will have an uncertainty relationship~\cite{Carcassi:2022bpm}
\begin{equation}
    \sigma_x \sigma_p \geq \Sigma(S).
\end{equation}
where $\sigma_x$ and $\sigma_p$ are the standard deviations of position and momentum and $\Sigma$ is the determinant of the covariance matrix and. The specific value of the uncertainty $\Sigma$ will depend on the entropy and on whether we are using classical or quantum mechanics, though the relationship will always be saturated by Gaussian states with no correlation between $x$ and $p$.

@@

Having given qualitative reasons for the limit, and having seen how, in specific cases, standard arguments can be reinterpreted consistently with our view, we now develop the limit as a direct consequence of entropy increase. Since the maximum entropy attainable by a quantum system is the logarithm of the dimension of the corresponding Hilbert space~\cite{nielsen2010quantum}, the high-entropy limit exists only for infinite-dimensional spaces. This explains why spin, which lives in a finite-dimensional space, is an intrinsically quantum property. Therefore we are going to concentrate on the case of a single degree of freedom identified by position and momentum.\footnote{A similar limit should exist for an infinite-dimensional directional degree of freedom (i.e. for angular momentum). Though technically not a limit but more of an approximation, for finite-dimensional spaces one could look at mixed state that are close to the maximally mixed states (i.e. $\rho = I/n + \epsilon \sigma$).} The case of multiple degrees of freedom can be recovered by increasing the entropy of all DOFs at the same time, that is, by increasing the entropy of each DOF independently, without introducing correlations.

@@

\begin{figure}[h]
    \centering
\begin{tikzpicture}
\begin{axis}[
height=7cm,
width=\linewidth*0.8,
grid=both,
grid style={line width=.1pt, draw=gray!25},
axis lines=middle,
xmin=0,
xlabel = \(\Sigma\),
ylabel = \(S\),
legend style={at={(0.95,0.25)},anchor=east},
]
\addplot[thick,blue,samples=150,domain=0.06:1.5] {ln(x)+1};
\addlegendentry{\(\text{classical}\)}
\addplot[thick,BrickRed,densely dotted,samples=150,domain=0.5:1.5] {(x+1/2)*ln(x+1/2)-(x-1/2)*ln(x-1/2)};
\addlegendentry{\(\text{quantum}\)}
\addplot[thick,ForestGreen,dashed,samples=150,domain=0.25:1.5] {((2*x+1/2)*ln(2*x+1/2)-(2*x-1/2)*ln(2*x-1/2) - ln(2)};
\addlegendentry{\(\text{$\lambda = 2$}\)}
\addplot[thick,violet,densely dashed,samples=150,domain=0.05:1.5] {((5*x+1/2)*ln(5*x+1/2)-(5*x-1/2)*ln(5*x-1/2) - ln(5)};
\addlegendentry{\(\text{$\lambda = 5$}\)}
\end{axis}
\end{tikzpicture}
    \caption{Entropy $S$ in nats for a Gaussian state as a function of uncertainty $\Sigma$, measured in units of $\hbar$. It shows the effect of the transition from classical to quantum by shrinking the pure states by a factor of $\lambda$ instead of stretching the mixed states by the same factor. The rescaled uncertainty for pure states becomes $\hbar/\lambda$ and the modified Gaussian bound as becomes $S_{\lambda Q}(\Sigma) = S_Q(\lambda \Sigma) - \ln(\lambda)$. Taking $\lambda$ to infinity corresponds to taking $\hbar/\lambda$ to zero.}
    \label{fig:uncertainty_scaled}
\end{figure}

@@

\begin{figure}[h]
	\centering
	\includegraphics[width=0.4\textwidth]{stretchedCat.png}
	\caption{A cat state stretched with different values of $\lambda$. The axes are stretched by $\sqrt{\lambda}$ as well. The color scale is reduced to match the range of the function, though white always corresponds to zero. While the positive and negative peaks have comparable magnitude initially, the stretching map dampens the negative peak much more quickly. Note that, consistently with Fig.~\ref{fig:uncertainty_scaled}, the negative values are severely dampened with $\lambda=5$.}
	\label{fig:cat_streatched}
\end{figure}

@@

We have seen that classical mechanics can be recovered as the high-entropy limit of quantum mechanics. That is, states of higher and higher entropy are better and better approximated by classical distributions over phase space and, if the entropy increase is done uniformly, unitary evolution can be approximated with classical Hamiltonian evolution. This approach to the classical limit is independent of mechanism and interpretation, as it does not matter how the entropy is increased or what one believes quantum states to represent: as long as the description is in terms of mixed states of sufficiently high entropy, classical mechanics applies. The approach fits naturally with experimental considerations and other approaches, such as decoherence, leading to an interpretation that is  thermodynamically meaningful, representation-independent, and stable under coarse-graining. It also recovers the established mathematical recipe, which is the group contraction for $\hbar \to 0$.\footnote{As pointed out by a reviewer, the $\hbar \to 0$ limit breaks down for chaotic or non-analytic systems, while the high-entropy limit will always exist. Therefore there may be an additional concrete advantage to our approach over the standard one.} Physically, this can be understood as taking the entropy of pure states to minus infinity, which is equivalent to saying that the relative entropy of mixed states goes to plus infinity. That is, in the same way that $c \to \infty$ should be understood as $v \ll c$, $\hbar \to 0$ should be understood as $S \gg 0$, as physically we cannot take limits of physical constants. This gives a reasonable and precise account of the classical limit: in the same way that non-relativistic mechanics applies to low speed, classical mechanics applies to high entropy.

>>>>

In both classical and quantum mechanics, Gaussian states maximize entropy for a single independent degree of freedom at a fixed uncertainty or, equivalently, minimize uncertainty at fixed entropy. This means that, if we fix the entropy $S$, we will have an uncertainty relationship~\cite{Carcassi:2022bpm}
\begin{equation}
    \sigma_x \sigma_p \geq \Sigma(S).
\end{equation}
where $\sigma_x$ and $\sigma_p$ are the standard deviations of position and momentum and $\Sigma$ is the determinant of the covariance matrix. The specific value of the uncertainty $\Sigma$ will depend on the entropy and on whether we are using classical or quantum mechanics, though the relationship will always be saturated by Gaussian states with no correlation between $x$ and $p$.

@@

Having given qualitative reasons for the limit, and having seen how, in specific cases, standard arguments can be reinterpreted consistently with our view, we now develop the limit as a direct consequence of entropy increase. Since the maximum entropy attainable by a quantum system is the logarithm of the dimension of the corresponding Hilbert space~\cite{nielsen2010quantum}, the high-entropy limit exists only for infinite-dimensional spaces. This explains why spin, which lives in a finite-dimensional space, is an intrinsically quantum property. Therefore we are going to concentrate on the case of a single degree of freedom identified by position and momentum.\footnote{A similar limit should exist for an infinite-dimensional directional degree of freedom (i.e. for angular momentum). Though technically not a limit but more of an approximation, for finite-dimensional spaces one could look at mixed states that are close to the maximally mixed states (i.e. $\rho = I/n + \epsilon \sigma$).} The case of multiple degrees of freedom can be recovered by increasing the entropy of all DOFs at the same time, that is, by increasing the entropy of each DOF independently, without introducing correlations.

@@

\begin{figure}[h]
    \centering
\begin{tikzpicture}
\begin{axis}[
height=7cm,
width=\linewidth*0.8,
grid=both,
grid style={line width=.1pt, draw=gray!25},
axis lines=middle,
xmin=0,
xlabel = \(\Sigma\),
ylabel = \(S\),
legend style={at={(0.95,0.25)},anchor=east},
]
\addplot[thick,blue,samples=150,domain=0.06:1.5] {ln(x)+1};
\addlegendentry{\(\text{classical}\)}
\addplot[thick,BrickRed,densely dotted,samples=150,domain=0.5:1.5] {(x+1/2)*ln(x+1/2)-(x-1/2)*ln(x-1/2)};
\addlegendentry{\(\text{quantum}\)}
\addplot[thick,ForestGreen,dashed,samples=150,domain=0.25:1.5] {((2*x+1/2)*ln(2*x+1/2)-(2*x-1/2)*ln(2*x-1/2) - ln(2)};
\addlegendentry{\(\text{$\lambda = 2$}\)}
\addplot[thick,violet,densely dashed,samples=150,domain=0.05:1.5] {((5*x+1/2)*ln(5*x+1/2)-(5*x-1/2)*ln(5*x-1/2) - ln(5)};
\addlegendentry{\(\text{$\lambda = 5$}\)}
\end{axis}
\end{tikzpicture}
    \caption{Entropy $S$ in nats for a Gaussian state as a function of uncertainty $\Sigma$, measured in units of $\hbar$. It shows the effect of the transition from classical to quantum by shrinking the pure states by a factor of $\lambda$ instead of stretching the mixed states by the same factor. The rescaled uncertainty for pure states becomes $\hbar/\lambda$ and the modified Gaussian bound becomes $S_{\lambda Q}(\Sigma) = S_Q(\lambda \Sigma) - \ln(\lambda)$. Taking $\lambda$ to infinity corresponds to taking $\hbar/\lambda$ to zero.}
    \label{fig:uncertainty_scaled}
\end{figure}

@@

\begin{figure}[h]
	\centering
	\includegraphics[width=\textwidth]{stretchedCat.png}
	\caption{A cat state stretched with different values of $\lambda$. The axes are stretched by $\sqrt{\lambda}$ as well. The color scale is reduced to match the range of the function, though white always corresponds to zero. While the positive and negative peaks have comparable magnitude initially, the stretching map dampens the negative peak much more quickly. Note that, consistently with Fig.~\ref{fig:uncertainty_scaled}, the negative values are severely dampened with $\lambda=5$.}
	\label{fig:cat_streatched}
\end{figure}

@@

We have seen that classical mechanics can be recovered as the high-entropy limit of quantum mechanics. That is, states of higher and higher entropy are better and better approximated by classical distributions over phase space and, if the entropy increase is done uniformly, unitary evolution can be approximated with classical Hamiltonian evolution. This approach to the classical limit is independent of mechanism and interpretation, as it does not matter how the entropy is increased or what one believes quantum states to represent: as long as the description is in terms of mixed states of sufficiently high entropy, classical mechanics applies. The approach fits naturally with experimental considerations and other approaches, such as decoherence, leading to an interpretation that is  thermodynamically meaningful, representation-independent, and stable under coarse-graining. It also recovers the established mathematical recipe, which is the group contraction for $\hbar \to 0$.\footnote{As pointed out by a reviewer, the $\hbar \to 0$ limit can be problematic for chaotic or non-analytic systems. Therefore there may be an additional concrete advantage to our approach over the standard one.} Physically, this can be understood as taking the entropy of pure states to minus infinity, which is equivalent to saying that the relative entropy of mixed states goes to plus infinity. That is, in the same way that $c \to \infty$ should be understood as $v \ll c$, $\hbar \to 0$ should be understood as $S \gg 0$, as physically we cannot take limits of physical constants. This gives a reasonable and precise account of the classical limit: in the same way that non-relativistic mechanics applies to low speed, classical mechanics applies to high entropy.
